Problem Set 6 (Answers)

### 2019-12-06 static/homeworks/06-problem-set-answers.Rmd
Add HW 6 empty markdown file, answers to #1-2 in HW 6 answers

@@

>>>>

<!--WRITE YOUR ANSWERS BELOW -->
Fixed effects can be used for panel data, where each observation$_{it}$ belongs to a group $i$ and a time period $t$. Running a simple OLS model as usual, known as a **pooled model** would be biased due to systematic relationships within and between groups and time periods that cause $X_{it}$ to be endogenous:

$$\widehat{Y_{it}}=\beta_0+\beta_1 X_{it}+u_{it}$$

Group-fixed effects $(\alpha_i$) pull out of the error term all of the factors that explain $Y_{it}$ that are stable over time within each individual group ($i$), but vary across groups. For example, if groups are U.S. states, state-fixed effects pull out all of the error term all of the idiosyncrasies of each state that do not change over time, such as cultural differences, geographical differences, historical differences, population differences, etc. Group fixed effects *do not* pull out factors that change over time, such as federal laws passed, recessions affecting all states, etc.

Time-fixed effects $(\theta_i)$ pull out of the error term all of the factors that explain $Y_{it}$ that change over time but do not vary across groups. For example, if groups are U.S. states, and time is in years, year-fixed effects pull out all of the error term all of the changes over the years that affect all U.S. states, such as recessions, inflation, national population growth, national immigration trends, or federal laws passed that affect all states. Time-fixed effects *do not* pull out factors that do not change over time.

$$\widehat{Y_{it}}=\beta_0+\beta_1 X_{it}+ \alpha_{i} + \theta_{t} + \epsilon_{it}$$

Mechanically, OLS estimates a separate constant for each group (and/or each time period), giving the expected value of $Y_{it}$ for that group or time-period. This can be done by either de-meaning the data and calculating OLS coefficients by exploiting the variation *within* each group and/or time period (which is why fixed effects estimators are called "within" estimators), or by creating a dummy variable for each group and/or time period (and omitting one of each, to avoid the dummy variable trap).

@@

<!--WRITE YOUR ANSWERS BELOW -->
$$\widehat{Y_{it}}=\beta_0+\beta_1 \text{Treated}_i +\beta_2 \text{After}_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+u_{it}$$

A difference-in-difference model compares the difference before and after a treatment between a group that receives the treatment and a group that does not. By doing so, we can isolate the effect of the treatment. It is easiest to see in the following equation:

$$\Delta \Delta Y = (\text{Treated}_{after}-\text{Treated}_{before})-(\text{Control}_{after}-\text{Control}_{before})$$

In OLS regression, we can simply make two dummy variables for observations$_it$ depending on the group and the time period each observation is:

$$\text{Treated}_i = \begin{cases}
        1 & \text{if } i \text{ receives treatment} \\
   			0 & \text{if } i \text{ does not receive treatment }\\
  		\end{cases}$$

$$\text{After}_t = \begin{cases}
    		1 & \text{if } t \text{is after treatment period} \\
   			0 & \text{if } t \text{is before treatment period}\\
  		\end{cases}$$

Lastly, we make an interaction term $\text{Treated}_i * \text{After}_t$ which isolates the treatment effect, captured by the coefficient on this term, $\beta_3$.

Diff-in-diff models assume a counterfactual that if the treated group did not receive treatment, it would have experience the same change as the control group over the pre- to post-treatment period.

A classic example is if a state(s) passes a law and others do not. We want to compare the difference in the difference before and after the law was passed between states that passed the law and states that did not:

$$\Delta_i \Delta_t Outcome = (\text{Passed}_{after}-\text{Passed}_{before})-(\text{Didn't}_{after}-\text{Didn't}_{before})$$

"Treatment$_i$" and time-fixed effects for "After$_t$" to estimate the treatment effect (still $\beta_3$): 
$$\widehat{Y_{it}}=\alpha_i +\theta_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+\epsilon_{it}$$

### 2019-12-06 static/homeworks/06-problem-set.Rmd
Add HW 6 empty markdown file, answers to #1-2 in HW 6 answers

<!--WRITE YOUR ANSWERS BELOW -->
Fixed effects can be used for panel data, where each observation$_{it}$ belongs to a group $i$ and a time period $t$. Running a simple OLS model as usual, known as a **pooled model** would be biased due to systematic relationships within and between groups and time periods that cause $X_{it}$ to be endogenous:

$$\widehat{Y_{it}}=\beta_0+\beta_1 X_{it}+u_{it}$$

Group-fixed effects $(\alpha_i$) pull out of the error term all of the factors that explain $Y_{it}$ that are stable over time within each individual group ($i$), but vary across groups. For example, if groups are U.S. states, state-fixed effects pull out all of the error term all of the idiosyncrasies of each state that do not change over time, such as cultural differences, geographical differences, historical differences, population differences, etc. Group fixed effects *do not* pull out factors that change over time, such as federal laws passed, recessions affecting all states, etc.

Time-fixed effects $(\theta_i)$ pull out of the error term all of the factors that explain $Y_{it}$ that change over time but do not vary across groups. For example, if groups are U.S. states, and time is in years, year-fixed effects pull out all of the error term all of the changes over the years that affect all U.S. states, such as recessions, inflation, national population growth, national immigration trends, or federal laws passed that affect all states. Time-fixed effects *do not* pull out factors that do not change over time.

$$\widehat{Y_{it}}=\beta_0+\beta_1 X_{it}+ \alpha_{i} + \theta_{t} + \epsilon_{it}$$

Mechanically, OLS estimates a separate constant for each group (and/or each time period), giving the expected value of $Y_{it}$ for that group or time-period. This can be done by either de-meaning the data and calculating OLS coefficients by exploiting the variation *within* each group and/or time period (which is why fixed effects estimators are called "within" estimators), or by creating a dummy variable for each group and/or time period (and omitting one of each, to avoid the dummy variable trap).

@@

<!--WRITE YOUR ANSWERS BELOW -->
$$\widehat{Y_{it}}=\beta_0+\beta_1 \text{Treated}_i +\beta_2 \text{After}_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+u_{it}$$

A difference-in-difference model compares the difference before and after a treatment between a group that receives the treatment and a group that does not. By doing so, we can isolate the effect of the treatment. It is easiest to see in the following equation:

$$\Delta \Delta Y = (\text{Treated}_{after}-\text{Treated}_{before})-(\text{Control}_{after}-\text{Control}_{before})$$

In OLS regression, we can simply make two dummy variables for observations$_it$ depending on the group and the time period each observation is:

$$\text{Treated}_i = \begin{cases}
        1 & \text{if } i \text{ receives treatment} \\
   			0 & \text{if } i \text{ does not receive treatment }\\
  		\end{cases}$$

$$\text{After}_t = \begin{cases}
    		1 & \text{if } t \text{is after treatment period} \\
   			0 & \text{if } t \text{is before treatment period}\\
  		\end{cases}$$

Lastly, we make an interaction term $\text{Treated}_i * \text{After}_t$ which isolates the treatment effect, captured by the coefficient on this term, $\beta_3$.

Diff-in-diff models assume a counterfactual that if the treated group did not receive treatment, it would have experience the same change as the control group over the pre- to post-treatment period.

A classic example is if a state(s) passes a law and others do not. We want to compare the difference in the difference before and after the law was passed between states that passed the law and states that did not:

$$\Delta_i \Delta_t Outcome = (\text{Passed}_{after}-\text{Passed}_{before})-(\text{Didn't}_{after}-\text{Didn't}_{before})$$

"Treatment$_i$" and time-fixed effects for "After$_t$" to estimate the treatment effect (still $\beta_3$): 
$$\widehat{Y_{it}}=\alpha_i +\theta_{t}+\beta_3 (\text{Treated}_i \times \text{After}_{t})+\epsilon_{it}$$

@@

If joining the Peace Corps is a substitute for private sector jobs, then as unemployment rises, so too should applications. Although, it could also be possible that people also are more willing to opt out of the workforce when the economy is strong, so we should examine this empirically to be sure.

@@

```{r}
library(tidyverse)

# load in data
peacecorps<-read_csv("https://metricsf19.classes.ryansafner.com/data/PeaceCorps.csv")
```

```{r}
# count number of states
peacecorps %>%
  count(state)

# count number of years
peacecorps %>%
  count(year)

# get number of distinct states and years
peacecorps %>%
  summarize(Num_states = n_distinct(state),
            Num_years = n_distinct(year))

# get distinct values of states
peacecorps %>%
  distinct(state)

# get distinct values of years
peacecorps %>%
  distinct(year)
```

@@

```{r}
# install.packages("plm")

# load plm
library(plm)

pdim(peacecorps, index=c("state", "year"))
```

@@

```{r, fig.retina=3}
ggplot(data = peacecorps)+
    aes(x = unemployrate,
        y = appspc)+
    geom_point(aes(color = as.factor(state)))+
    geom_smooth(method = "lm")+
  scale_x_continuous(breaks=seq(0,20,2),
                     labels = function(x){paste(x,"%", sep="")})+
  labs(x = "Unemployment Rate",
       y = "Peace Corps Applications (per capita)")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=18)+
  theme(legend.position = "none") # remove legend for State colors
```

We can see there are very clear outliers at the top. Let's plot `text` instead of `point`s, using the `stateshort` to see which observations are which states:

```{r, fig.retina=3}
ggplot(data = peacecorps)+
    aes(x = unemployrate,
        y = appspc)+
    geom_text(aes(color = as.factor(state), label = stateshort))+ #<<
    geom_smooth(method = "lm")+
  scale_x_continuous(breaks=seq(0,20,2),
                     labels = function(x){paste(x,"%", sep="")})+
  labs(x = "Unemployment Rate",
       y = "Peace Corps Applications (per capita)")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=18)+
  theme(legend.position = "none") # remove legend for State colors
```

Clearly `DIS`, which is District of Columbia, are the outliers. Let's make a second scatterplot without them:

```{r, fig.retina=3}
peacecorps %>%
  filter(state != "DISTRICT OF COLUMBIA") %>%
ggplot(data = .)+
    aes(x = unemployrate,
        y = appspc)+
    geom_point(aes(color = as.factor(state)))+
    geom_smooth(method = "lm")+
  scale_x_continuous(breaks=seq(0,20,2),
                     labels = function(x){paste(x,"%", sep="")})+
  labs(x = "Unemployment Rate",
       y = "Peace Corps Applications (per capita)")+
  theme_classic(base_family = "Fira Sans Condensed",
           base_size=18)+
  theme(legend.position = "none") # remove legend for State colors
```

@@

```{r}
pooled<-lm(appspc ~ unemployrate, data = peacecorps)
summary(pooled)
```

$$\widehat{\text{Apps per capita}_{it}}=46.23+0.84 \, \text{Unemployment Rate}_{it}$$

For every 1 percentage point increase in unemployment, there are 0.46 more applications to the Peace Corps per capita.

```{r}
peacecorps_no_outliers<-peacecorps %>%
  filter(state != "DISTRICT OF COLUMBIA")

pooled_no_outliers<-lm(appspc ~ unemployrate, data = peacecorps_no_outliers)
summary(pooled_no_outliers)
```

$$\widehat{\text{Apps per capita}_{it}}=48.59-0.37 \, \text{Unemployment Rate}_{it}$$

For every 1 percentage point increase in unemployment, there are 0.37 *fewer* applications to the Peace Corps per capita.

The coefficient changes signs and becomes a smaller magnitude by dropping the outliers (DC)!

@@

```{r, fig.retina=3}
fe_reg<-lm(appspc ~ unemployrate + factor(state), data = peacecorps)
summary(fe_reg)
```

For every 1 percentage point increase in the unemployment rate, there are 0.66 more applications to the Peace Corps per capita. It remained positive, and shrank in size, from the original regression (with outliers).

@@

This is just like any model with dummy variables representing different categories. The constant (12.61) represents the *reference category* not put in the regression (to avoid the dummy variable trap), i.e. `ALABAMA.`

The coefficient on Maryland is 45.19. This means that, on average, there are 45.19 *more* applications per capita in Maryland *than Alabama* over the time period in the data. To get the average for Maryland, we add $12.61+45.19=57.80$.

```{r}
# if we want to extract it using R
library(broom)

fe_reg_tidy<-tidy(fe_reg)

AL<-fe_reg_tidy %>%
  filter(term == "(Intercept)") %>%
  pull(estimate)

MD_AL_diff<-fe_reg_tidy %>%
  filter(term == "factor(state)MARYLAND") %>%
  pull(estimate)

AL+MD_AL_diff
```

@@

```{r}
fe_reg_alt<-plm(appspc ~ unemployrate, index = "state", model = "within", data = peacecorps)
summary(fe_reg_alt)
```

@@

```{r}
fe_2way_reg<-lm(appspc ~ unemployrate + factor(state) + factor(year), data = peacecorps)
summary(fe_2way_reg)
```

For every 1 percentage point increase in the unemployment rate, there are 0.71 applications per capita to the Peace Corps.

@@

$$\begin{align*}
\widehat{\text{Apps per capita}_{it}}&=14.49 +0.71 \, \text{Unemployment Rate}_{it} + 45.24 \, \text{Maryland_i} -8.89 \text{2007}_t \\
\widehat{\text{Apps per capita}_{MD, 2011}}&=14.49 +0.71 \, (5) + 45.24 \, (1) -8.89 (1) \\
\widehat{\text{Apps per capita}_{MD, 2011}}& \approx 54.09
\end{align*}$$

```{r}
# if we want to extract it using R
fe_2way_reg<-tidy(fe_2way_reg)

constant<-fe_2way_reg %>%
  filter(term == "(Intercept)") %>%
  pull(estimate)

unemploy_me<-fe_2way_reg %>%
  filter(term == "unemployrate") %>%
  pull(estimate)

MD_dif<-fe_2way_reg %>%
  filter(term == "factor(state)MARYLAND") %>%
  pull(estimate)

dif_2007<-fe_2way_reg %>%
  filter(term == "factor(year)2007") %>%
  pull(estimate)

constant+unemploy_me*5+MD_dif+dif_2007

# some rounding error
```

@@

```{r}
fe_2way_reg_alt<-plm(appspc ~ unemployrate, index = c("state", "year"), data = peacecorps, effect = "twoways")
summary(fe_2way_reg_alt)
```

@@

Yes. There may be other time-varying variables that are omitted and correlated with unemployment, but not picked up in the time-fixed effect since they are not the same in each state. Examples could be a local market bubble some years in Texas or Nevada, but not in Maine, or bad weather in Florida one year, but not in Wyoming, etc.^[It's better to use the `plm()`-generated regressions so as to avoid the multitude of coefficients for the state and year dummies! You certainly could use the dummy variable ones and manually list all of the variables to suppress in the table inside `omit_coefs()`...]

@@

<!--WRITE YOUR ANSWERS BELOW -->
```{r}
library(huxtable)
huxreg(pooled,
       pooled_no_outliers,
       fe_reg_alt,
       fe_2way_reg_alt,
         coefs = c("Constant" = "(Intercept)",
                 "Unemployment Rate" = "unemployrate"),
       statistics = c("N" = "nobs",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       note = NULL, # suppress footnote for stars, to insert Fixed Effects Row below
       number_format = 3) %>%

add_rows(c("Outliers", "Yes", "No DC", "Yes", "Yes"), # add row for outliers
         after = 5) %>% # insert after 5th row
add_rows(c("Fixed Effects", "None", "None", "State", "State and Year"), # add fixed effects row
         after = 6) # insert after 6th row

```

@@

```{r}
# load in data
texas<-read_csv("https://metricsf19.classes.ryansafner.com/data/TexasSchools.csv")
```

```{r, fig.retina=3}
pooled_reg<-lm(LnAvgSalary~OnCycle, data = texas)
summary(pooled_reg)
```

$$\widehat{\text{log(Average Salary})_it} = 10.67-0.03 \, \text{OnCycle}_{it}$$

Recognize this is a linear-log model and $X$ is a dummy variable. School board elections that are held `OnCycle` $(=1)$ lead to $\hat{\beta_1 \times 100\% = −0.03 \times 100\% = −3\%$ lower teacher salaries. This is highly statistically significant.

This is almost certainly biased. There are probably things correlated at the district level between both whether or not districts vote (or switch to voting) on cycle/off cycle and with teacher salaries. Perhaps some districts have strong/weak teachers unions and that determines whether they choose to vote on or off cycle (off cycle would be preferable, and perhaps districts with strong teachers unions kept the elections off cycle).

Also, there could be changes in the entire State over time-perhaps all Texas teacher salaries were increasing or decreasing over the years.

@@

```{r}
# create a variable, that is a factor, using ifelse() function
# # if the year is greater than 2006, mark "After" as 1,
# # If the year is NOT greater than 2006, market "After" a 0
texas<-texas %>%
  mutate(After = factor(ifelse(test = Year>2006,
                        yes = 1,
                        no = 0)))
head(texas)
```

@@

```{r}
dnd<-lm(LnAvgSalary ~ CycleSwitch + After + CycleSwitch:After, data = texas)
summary(dnd)
```

$$\widehat{\text{Log(Average Salary)}_{it}}=10.6711-0.0240 \, \text{Cycle Switch}_{i} + 0.0093 \, \text{After}_t - 0.0086 \, (\text{Cycle Switch}_i \times \text{After}_t)$$

@@

- $\hat{\beta_0}$: districts before 2007 that did not switch had `ln(AvgSalary)` of 10.6711
- $\hat{\beta_1}$: districts before 2007 that did switch had `ln(AvgSalary)` 0.0240 lower
than districts that did not switch
- $\hat{\beta_2}$: all districts after the change had `ln(AvgSalary)` 0.0093 higher than before the change
- $\hat{\beta_3}$: districts that made a switch had a 0.0086 *lower* change in `ln(AvgSalary)` over time than districts that did not switch

@@

- i. Before change and did not switch: $\hat{\beta_0} = 10.6711$
- ii. After change and did not switch: $\hat{\beta_0}+\hat{\beta_2} = 10.6711 + 0.0093 = 10.6804$
- iii. Before change and did switch: $\hat{\beta_0}+\hat{\beta_1} = 10.6711 − 0.0240 = 10.6471$
- iv. After change and did switch: $\hat{\beta_0}+\hat{\beta_1}+\hat{\beta_2}+\hat{\beta_3} = 10.6711−0.0240+ 0.0093 − 0.0086 = 10.6478$

@@

```{r}
# avg log salary for districts BEFORE that did NOT switch
texas %>%
  filter(CycleSwitch == 0,
         After == 0) %>%
  summarize(average = mean(LnAvgSalary, na.rm=T)) # need to remove NA's

# avg log salary for districts AFTER that did NOT switch
texas %>%
  filter(CycleSwitch == 0,
         After == 1) %>%
  summarize(average = mean(LnAvgSalary, na.rm=T)) # need to remove NA's

# avg log salary for districts BEFORE that DID switch
texas %>%
  filter(CycleSwitch == 1,
         After == 0) %>%
  summarize(average = mean(LnAvgSalary, na.rm=T)) # need to remove NA's

# avg log salary for districts AFTER that DID switch
texas %>%
  filter(CycleSwitch == 1,
         After == 1) %>%
  summarize(average = mean(LnAvgSalary, na.rm=T)) # need to remove NA's
```

@@

$$\begin{align*}
		\Delta \Delta ln(AvgSalary)&=(Switched_{after}-Switched_{before})-(Didn't_{after}-Didn't_{before})\\
		&=(10.6478-10.6471)-(10.6804-10.6711)\\
		&=(0.0007)-(0.0093)\\
		&=-0.0086\\ \end{align*}$$

This is $\beta_3$.

@@

We can see that the districts that switched paid 2.3% less to their teachers to begin with. $\beta_2$ shows the difference, before treatment, between the treated (districts that switched) and control groups (districts that did not switch).

We can see that all districts saw a 0.93% increase in teacher salaries after the years after the switch. $\beta_3$ shows the difference, for both the treatment and control, of salaries over time (before and after the switch).

@@

```{r}
# make sure DistNumber and Year are factors for plm
texas<-texas %>%
  mutate_at(c("DistNumber", "Year"), factor)
head(texas) # make sure it worked

dnd_fe_reg<-plm(LnAvgSalary ~ CycleSwitch*After, index=c("DistNumber", "Year"), effect = "twoways", data = texas)
summary(dnd_fe_reg)
```

Districts that switched to On Cycle saw teacher salaries decrease by 8.67%. No change from our original model!

@@

```{r}
huxreg(pooled_reg,
       dnd,
       dnd_fe_reg,
         coefs = c("Constant" = "(Intercept)",
                 "On Cycle" = "OnCycle",
                 "Switched" = "CycleSwitch",
                 "After Switch" = "After1",
                 "Interaction" = "CycleSwitch:After1"),
       statistics = c("N" = "nobs",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       note = NULL, # suppress footnote for stars, to insert Fixed Effects Row below
       number_format = 3) %>%
add_rows(c("Fixed Effects", "None", "None", "District and Year"), # add fixed effects row
         after = 11) # insert after 11th row
```

>>>>

@@

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```

@@

```{r}
# Write your code here
```
